# Schedule

This is the schedule for the course. 
Please note that it is liable to change (possibly at short notice). 
If you find a mismatch between this schedule and the official one^[https://mitsdu.sdu.dk/skema/activity/N110040101/e22], then it is the official one that is correct.

The columns, GSWR and Course Book, refer to the relevant chapters in the recommended text book ("*Getting Started With R*") and this course book, respectively. You should aim to read and work through these chapters as the course proceeds.

The names of the topics for the **Practical** sessions corresponds to the chapter names in this website/book.

```{r, echo = FALSE, include=knitr::is_latex_output()}
plot(1, 1, type = "n", axes = FALSE, xlab = "", ylab = "")
text(1, 1, labels = "The schedule is only available on the HTML version of this document")
```

```{r echo = FALSE}
updateSchedule <- FALSE
```

```{r, echo = FALSE, message=FALSE,warning=FALSE,eval=knitr::is_html_output()}
library(flextable)
library(readxl)
library(lubridate)
library(magrittr)
library(dplyr)
library(calendar)

system("cp -r ~/Dropbox/_SDU_Teaching/BB852/'Course Documents'/BB852_Schedule.xlsx ~/Dropbox/_SDU_Teaching/BB852/BB852_Book/BB852_Schedule.xlsx")
schedule <- readxl::read_excel("BB852_Schedule.xlsx")
```

```{r, echo = FALSE, message=FALSE, warning=FALSE, results = 'hide', eval = all(knitr::is_html_output(),updateSchedule)}
# Clean up old ics files
ics_files <- list.files(pattern = "*.ics")
file.remove(ics_files)

curl::curl_download("https://mitsdu.sdu.dk/skema/icalendar/activities/N110040101/e22/e22", destfile = "bb852.ics")
```


```{r, echo = FALSE, message=FALSE, warning=FALSE, results = 'hide'}

# Last sunday in October, 0300, CEST -> CET,
# So the time offset moves from +2 to +1

savingsTimeSwitch <- with_tz(lubridate::as_datetime("2021-10-31 03:00:00"), "Europe/Copenhagen")


xx <- readLines("bb852.ics")

locationIndex <- grep("^LOCATION", xx)
location <- gsub(pattern = "LOCATION:Odense ", replacement = "", x = xx[locationIndex])



courseCal <- ical::ical_parse_df("bb852.ics") %>%
  filter(!is.na(uid)) %>%
  arrange(start) %>%
  select(summary, DTSTART = start, DTEND = end) %>%
  mutate(Session = row_number()) %>%
  mutate(savingsT = DTSTART < savingsTimeSwitch) %>%
  # mutate(offsetT = if_else(savingsT == TRUE, 2, 1)) %>% #Check this every year!
  mutate(offsetT = 0) %>% # Check this every year!
  mutate(DTSTART2 = DTSTART + hours(offsetT)) %>%
  mutate(DTEND2 = DTEND + hours(offsetT)) %>%
  mutate(St = hour(DTSTART2)) %>%
  mutate(En = hour(DTEND2)) %>%
  mutate(wd = lubridate::wday(DTSTART2, label = TRUE)) %>%
  # mutate(Room = gsub(pattern = "Odense ",replacement = "", x = LOCATION)) %>%
  mutate(Room = location) %>%
  mutate(Room = paste0("(", Room, ")")) %>%
  mutate(Date = paste0(month(DTSTART2, label = TRUE), " ", day(DTSTART2), ", ", wd)) %>%
  mutate(Time = paste0("kl.", St, "-", En))

courseCal_web <- courseCal %>%
  select(Session, Date, Time, Room)



schedule <- readxl::read_excel("BB852_Schedule.xlsx")

schedule <- left_join(courseCal_web, schedule)
```


```{r, echo = FALSE, message=FALSE,eval=knitr::is_html_output(), warning=FALSE}
library(lubridate)
library(officer)

schedule2 <- schedule %>%
  select(Part, Date, Time, Room, Session, Type, Topic, GSWR)

big_b <- fp_border(color = "gray70", width = 3)
std_b <- fp_border(color = "gray70", style = "dashed")
big_b_end <- fp_border(color = "white", width = 3)

ft <- flextable(schedule2, cwidth = c(2, 6, 1, 1, .5, 1, 4, .75, .75)) %>%
  bold(part = "header") %>%
  bold(j = 1) %>%
  color(
    i = ~ Type %in% c("Lecture", "Exam"),
    j = ~ Session + Type + Topic,
    color = "red"
  ) %>%
  merge_v(j = ~ Date + Part) %>%
  fontsize(part = "all", size = 14) %>%
  vline(border = std_b, j = 1:4) %>%
  rotate(j = ~Part, align = "top", rotation = "tblr") %>%
  hline(border = std_b, i = seq(2, nrow(schedule2), 2)) %>%
  hline(border = big_b_end, i = nrow(schedule2))


ft
```

```{r, echo = FALSE, message=FALSE,warning=FALSE, results = 'hide'}
source("personalCalendar.R")
```
