# An R refresher

**Learning goals**

By the end of this chapter you should be able to:

- Use RStudio to run code from scripts.
- Create and inspect basic objects in R.
- Work with vectors and data frames.
- Import a CSV file using a relative path.
- Make simple plots and summaries.

**Prerequisites**

- You have R and RStudio installed and have created the course project described in the previous chapter.

**A tiny motivating example**

Let’s make a tiny plot so we know everything is working.

```{r 0-03-An-R-Refresher-1, message=FALSE}

body_length <- c(4.2, 5.1, 5.6, 6.8, 7.4)
body_mass <- c(9.5, 12.1, 14.0, 18.2, 21.5)

plot(body_length, body_mass)
```

**Try this:** Change one value and re-run. What changes on the plot?

---

## Getting started with RStudio

**What we’re about to do:** We will set up a simple workflow so your work is reproducible.

RStudio is not R itself. It is the interface where you write scripts, run code, and view results.

```{block 0-03-An-R-Refresher-2, type="do-something"}
**Do this:** Open RStudio and create a new script (`File > New File > R Script`). Save it as `practice.R`.
```

**Why this matters:** Scripts let you save, revisit, and debug your work. Copy-pasting into the console will not scale.

**Common mistake:** Running everything in the console and losing your work. If this happens, start a script and re-run your commands there.

**Takeaway:** Work in scripts, not just the console.

---

## Getting help

**What we’re about to do:** Learn how to ask R for help.

```{r 0-03-An-R-Refresher-3, eval=FALSE}
?rep
help.search("bar plot")
??"bar plot"
```

**What just happened:** `?` opens a help page when you know a function name. `help.search()` and its shortcut, `??`, search across help pages when you do not.

**Try this:** Look up `mean` or `plot` and read the arguments section.

**Takeaway:** R has built-in help for almost everything.

---

## R as a calculator

**What we’re about to do:** Use R for basic arithmetic and logic.

```{r 0-03-An-R-Refresher-4}
4 + 3
9 - 12
6 / 3
7 * 3
(2 * 7) + 2 - 0.4
```

```{r 0-03-An-R-Refresher-5}
sqrt(945)
3^5
abs(-23.4)
round(2.35425, digits = 2)
log(1.2)
exp(1)
log10(6)
```

Logical comparisons:

```{r 0-03-An-R-Refresher-6}
3 < 10
5 > 7
5 == 5
6 != 5
3 %in% c(1, 2, 3, 4, 5)
6 %in% c(1, 2, 3, 4, 5)
```

**Takeaway:** R handles arithmetic and logical tests directly.

---

## Objects and vectors

**What we’re about to do:** Create objects and simple vectors.

```{r 0-03-An-R-Refresher-7}
x <- 4
x
```

The assignment operator `<-` stores the value on its right under the name on its left. You can read this as "`x` gets 4". Typing the object's name prints its value.

Vectors are sequences of values of the same basic type, such as numbers or text.

```{r 0-03-An-R-Refresher-8}
A <- 1:5
B <- c(1, 3, 6, 1, 7, 9)
C <- seq(from = 1, to = 12, by = 2)
D <- seq(from = 1, to = 5, by = 0.1)
E <- rep(c("Male", "Female"), each = 3)
F <- rep(c("Male", "Female"), times = c(2, 4))

A
C
E
F
```

```{block 0-03-An-R-Refresher-9, type="do-something"}
Try changing `seq()` or `rep()` arguments. What changes?
```

**Common mistake:** Forgetting `c()` when creating a vector of values.

**Takeaway:** Vectors are the building blocks of most data in R.

---

## Manipulating vectors

**What we’re about to do:** Do arithmetic on vectors.

```{r 0-03-An-R-Refresher-10}
B
B * 3
B - 2
c(A, B)
B * c(1, 2, 3, 4, 5, 6)
A / B
```

**What just happened:** R works element-by-element when vector lengths match. In the final calculation, `A` has five values but `B` has six. R recycles the shorter vector and produces a warning. Do not ignore this warning: unequal vector lengths usually indicate a mistake.

**Takeaway:** Vectorised operations are fast and convenient.

---

## Missing values, infinity, and NaN

**What we’re about to do:** See how R handles missing values and edge cases.

```{r 0-03-An-R-Refresher-11}
mean(c(1, 3, 6, 1, 7, 9, NA))
mean(c(1, 3, 6, 1, 7, 9, NA), na.rm = TRUE)
```

The first result is `NA` because one value is missing. The argument `na.rm = TRUE` tells `mean()` to remove missing values before doing the calculation.

```{r 0-03-An-R-Refresher-12}
5 / 0
-4 / 0
0 / 0
Inf - Inf
```

`Inf` and `-Inf` represent positive and negative infinity. `NaN` means "not a number" and appears when a calculation has no defined numerical result. `NA` instead represents a missing value.

**Takeaway:** Use `na.rm = TRUE` when you deliberately want a function to ignore missing values, and pay attention to non-finite results such as `Inf` and `NaN`.

---

## Data frames

**What we’re about to do:** Create and inspect a data frame.

```{r 0-03-An-R-Refresher-13}
height <- c(173, 145, 187, 155, 179, 133)
sex <- c("Male", "Female", "Male", "Female", "Male", "Female")
age <- c(17, 22, 32, 20, 27, 30)

mydata <- data.frame(height = height, age = age, sex = sex)
mydata
```

Inspect a data frame:

```{r 0-03-An-R-Refresher-14}
summary(mydata)
str(mydata)
```

Subset rows/columns:

```{r 0-03-An-R-Refresher-15}
mydata[1, ]
mydata[, 2]
mydata[1, 2]
subset(mydata, sex == "Female")
```

**Takeaway:** Data frames are tables: rows are observations, columns are variables.

---

## Classes and factors

**What we’re about to do:** Check classes and use factors for categories.

```{r 0-03-An-R-Refresher-16}
class(height)
class(sex)
class(mydata)
```

Factors tell R that a variable represents categories and record the possible category levels. Some plotting and modelling functions use this information:

```{r 0-03-An-R-Refresher-17, eval = FALSE}
plot(mydata$sex, mydata$age)
```

The command above fails because `mydata$sex` contains character data. Converting it to a factor tells `plot()` to compare the numerical values across categories:

```{r 0-03-An-R-Refresher-18}
plot(as.factor(mydata$sex), mydata$age)
```

```{r 0-03-An-R-Refresher-19}
mydata$sex <- as.factor(mydata$sex)
str(mydata)
```

**Common mistake:** Leaving a categorical variable as character data when a function expects a factor.

**Takeaway:** Use factors for categorical variables when needed.

---

## Organising your work and importing data

**What we’re about to do:** Import a CSV file, and meet the two functions you will see for doing this.

Keep your data in a `CourseData` folder inside your project. We will use **relative paths** throughout this course. If you do not have `CourseData`, download it from the course site.

There are two common ways to read a CSV file into R, and you will come across both:

- **`read.csv()`** comes with base R. It returns a `data.frame` and, with `stringsAsFactors = TRUE`, turns text columns into **factors** — which is convenient for the statistics later in the course.
- **`read_csv()`** comes from the `readr` package (part of the tidyverse, which we cover later). It is a little faster, prints a short summary of the column types it guessed, and returns a *tibble* (a modern data frame). It leaves text columns as character.

Both work well and you will see both in other people's code. **To keep things consistent, this book uses `read.csv()` from here on.**

```{r 0-03-An-R-Refresher-20}
carni <- read.csv("CourseData/carnivora.csv", stringsAsFactors = TRUE)
```

**What just happened:** `read.csv()` imported the data as a `data.frame` and, thanks to `stringsAsFactors = TRUE`, stored the text columns as factors.

**Try this:** Run `str(carni)` and compare it to `summary(carni)`.

**Common mistake:** File not found. Check spelling and folder names.

**Takeaway:** This book uses `read.csv()`; `read_csv()` is the tidyverse alternative you will also encounter.

---

## Inspecting the data

```{r 0-03-An-R-Refresher-21}
summary(carni)
dim(carni)
names(carni)
```

The dataset includes life history variables. For example:

- `FW` = Female body mass (kg)
- `GL` = Gestation length (days)
- `BW` = Birth mass (g)

```{r 0-03-An-R-Refresher-22}
summary(carni$FW)
```

---

## Tables and summary statistics

```{r 0-03-An-R-Refresher-23}
class(carni$Family)
levels(carni$Family)
```

```{r 0-03-An-R-Refresher-24}
table(carni$Family)
tapply(carni$FW, carni$Family, mean, na.rm = TRUE)
```

**Takeaway:** Tables and summaries give you a quick sense of patterns before you plot.

---

## Basic plotting

```{r 0-03-An-R-Refresher-25, fig.cap='A simple scatter plot', fig.show='asis'}
plot(
  log(carni$FW), log(carni$GL),
  xlab = "Log female body mass (kg)",
  ylab = "Log gestation length (days)"
)
```

**Interpretation:** Species with larger female body mass tend to have longer gestation lengths.

- **What to look for:**
  - Is the relationship roughly linear?
  - Are there any clear outliers?

**Try this:** Plot `carni$FW` against `carni$GL` without the log transformations and compare the shape.


---

## Exercise: Californian bird diversity

Surveys of bird abundance were carried out near Oakland, California.^[Vale, T. R., & Vale, G. R. (1976). Suburban bird populations in west-central California. *Journal of Biogeography*, 3(2), 157–165.] The locations were developed in different years, which lets us ask how species richness varies with the age of a suburb.

### The data

The file is `suburbanBirds.csv`. It contains the columns `Name`, `Year`, `HabitatIndex`, `nIndividuals`, and `nSpecies`. The surveys were conducted in 1975, so subtracting `Year` from 1975 gives the approximate age of each suburb at the time of the survey.

### Your tasks

**Level 1 (warm-up)**

1. Import the data as an object called `birds`, then check the columns with `str()`.
   **Checkpoint:** You should see the five columns listed above.

2. Calculate the mean, minimum, and maximum of `nSpecies`.
   **Checkpoint:** The values should be approximately 9.65, 3, and 15, respectively.

**Level 2 (practice)**

3. Create a new column with `birds$Age <- 1975 - birds$Year` and check that it looks sensible.
   **Checkpoint:** Younger suburbs have smaller `Age` values.

4. Plot `Age` against `nSpecies` with `plot(birds$Age, birds$nSpecies)`.
   **Checkpoint:** The x-axis is age and the y-axis is species richness.

**Level 3 (extension)**

5. Immediately after creating the scatter plot, add a smoother with `lines(lowess(birds$Age, birds$nSpecies))`. Does it change your interpretation?

**Hint:** If the plotting code fails, check the column names with `names(birds)`.

---

## Key takeaways

- Use scripts for reproducible work.
- Vectors and data frames are the core data types.
- `read.csv()` imports CSV files; `read_csv()` is the tidyverse alternative.
- Check classes with `str()` and convert categorical variables to factors when needed.
- Always inspect your data before analysing it.

## Common pitfalls recap

- Running code only in the console.
- File paths that point to the wrong folder.
- Treating categories as numbers or characters when they should be factors.
