# Tips and tricks

**Learning goals**

By the end of this chapter you should be able to:

- Adjust RStudio so it is comfortable to use.
- Use a few high‑value keyboard shortcuts.
- Export plots and tables cleanly.
- Avoid common data‑import problems.

**Prerequisites**

- You can run code from a script in RStudio.

**A tiny motivating example**

Here is a small quality‑of‑life improvement you can apply immediately:

```{r 0-04-Tips-and-Tricks-1, eval = FALSE}
options(scipen = 999)
```

This strongly discourages R from using scientific notation when printing numbers.

---

## Appearance

**What we’re about to do:** Make RStudio easier to read.

You can change the appearance under **Tools > Global Options > Appearance** (Windows and Linux) or **RStudio > Preferences > Appearance** (macOS). There you can change the theme and font.

- Use a **monospace** font (e.g., *Source Code Pro*, *Menlo*, or *Consolas*).
- Increase font size if you are squinting.

**Takeaway:** A readable setup reduces mistakes and fatigue.

---

## Shortcuts (top six)

**What we’re about to do:** Learn the shortcuts you will use every session.

- Run current line/selection: `Ctrl+Enter` (macOS: `Cmd+Enter`)
- Run from start to current line: `Ctrl+Alt+B` (macOS: `Cmd+Option+B`)
- Comment/uncomment: `Ctrl+Shift+C` (macOS: `Cmd+Shift+C`)
- Find in files: `Ctrl+Shift+F` (macOS: `Cmd+Shift+F`)
- Auto-complete: start typing, then press `Tab`
- Insert the pipe operator `|>`: `Ctrl+Shift+M` (macOS: `Cmd+Shift+M`)

This course uses R's native pipe, `|>`. You may also encounter the older `%>%` pipe in other code. For everything we do in this course, you can treat them as equivalent. Their differences matter only in more specialised code. If the shortcut inserts `%>%`, change the pipe setting under **Tools > Global Options > Code > Editing** (Windows and Linux) or **RStudio > Preferences > Code > Editing** (macOS).

**Try this:** Use `Ctrl+Shift+C` to comment out a block, then toggle it back.

**Takeaway:** Shortcuts save time and reduce clicking.

---

## Code style (keep it readable)

**What we’re about to do:** Make your scripts easier to read.

A few habits help a lot:

- One pipe step per line.
- Clear, meaningful object names.
- Blank lines between sections.

If you want a full guide, the tidyverse style guide is excellent: <https://style.tidyverse.org>.

**Takeaway:** Readable code is easier to debug and reuse.

---

## The plots pane

**What we’re about to do:** Save plots cleanly.

Plots appear in the **Plots** pane (usually bottom right). Use the back/forward arrows to browse old plots and **Export** to save one. In the export window, choose suitable dimensions and check the preview before saving.

**Common mistake:** Exporting a plot without checking its dimensions. A plot that looks fine in the pane may become cramped when saved at a different size.

---

## Tables

If you need to use a summary table outside R, save it as a CSV file and open it in spreadsheet software such as Excel.

```{r 0-04-Tips-and-Tricks-2, eval = FALSE}
write.csv(my_table, file = "example.csv", row.names = FALSE)
```

---

## Importing data: common problems

**What we’re about to do:** Avoid the most common import errors.

- **Everything in one column?** You probably used the wrong delimiter.
- **Unexpected symbols?** The file may use a different text encoding. Inspect it in a text editor and check how it was saved.

**Sanity checks:**

- `str(data)` to check column types
- `names(data)` to check column names
- `nrow(data)` and `ncol(data)` to check size

**Takeaway:** Always inspect data right after importing.

---

## Key takeaways

- Make RStudio comfortable and readable.
- Learn a few shortcuts and use them daily.
- Use `Ctrl+Shift+M` (macOS: `Cmd+Shift+M`) to insert `|>`.
- Export plots and tables deliberately.
- Always check imported data.

## Common pitfalls recap

- Forgetting to check the delimiter when importing.
- Using non‑monospace fonts that misalign code.
- Skipping basic checks after import.
