# Tips and tricks

**Learning goals**

By the end of this chapter you should be able to:

- Adjust RStudio so it is comfortable to use.
- Use a few high‑value keyboard shortcuts.
- Export plots and tables cleanly.
- Avoid common data‑import problems.

**Prerequisites**

- You can run code from a script in RStudio.

**A tiny motivating example**

Here is a small quality‑of‑life improvement you can apply immediately:

```{r 0-04-Tips-and-Tricks-1, eval = FALSE}
options(scipen = 999)
```

This tells R to avoid scientific notation when printing numbers.

---

## Appearance

**What we’re about to do:** Make RStudio easier to read.

You can change the appearance in **Tools > Options** (Windows) or **RStudio > Preferences** (Mac). The **Appearance** tab lets you change theme and font.

- Use a **monospace** font (e.g., *Menlo*, *Courier*, *Lucida Console*).
- Increase font size if you are squinting.

**Takeaway:** A readable setup reduces mistakes and fatigue.

---

## Shortcuts (top six)

**What we’re about to do:** Learn the shortcuts you will use every session.

- Run current line/selection: `Ctrl+Enter` (Mac: `Cmd+Enter`)
- Run from start to current line: `Ctrl+Alt+B` (Mac: `Cmd+Option+B`)
- Comment/uncomment: `Ctrl+Shift+C`
- Find in files: `Ctrl+Shift+F`
- Auto‑complete: type, then use arrows + Enter
- Insert pipe `%>%`: `Ctrl+Shift+M` (Mac: `Cmd+Shift+M`)

**Try this:** Use `Ctrl+Shift+C` to comment out a block, then toggle it back.

**Takeaway:** Shortcuts save time and reduce clicking.

---

## Code style (keep it readable)

**What we’re about to do:** Make your scripts easier to read.

A few habits help a lot:

- One pipe step per line.
- Clear, meaningful object names.
- Blank lines between sections.

If you want a full guide, the tidyverse style guide is excellent: <https://style.tidyverse.org>.

**Takeaway:** Readable code is easier to debug and reuse.

---

## The plots pane

**What we’re about to do:** Save plots cleanly.

Plots appear in the **Plots** pane (usually bottom right). Use the back/forward arrows to browse old plots. Use **Export** to save a plot.

**Common mistake:** Exporting before resizing. If your plot looks cramped, resize the plot pane first.

---

## Tables

If you create a summary table, save it to CSV and open in Excel or Word.

```{r 0-04-Tips-and-Tricks-2, eval = FALSE}
write.csv(my_table, file = "table.csv", row.names = FALSE)
```

---

## Importing data: common problems

**What we’re about to do:** Avoid the most common import errors.

- **Everything in one column?** You probably used the wrong delimiter.
- **Weird symbols?** Open the file in a text editor and inspect it.

**Sanity checks:**

- `str(data)` to check column types
- `names(data)` to check column names
- `nrow(data)` and `ncol(data)` to check size

**Takeaway:** Always inspect data right after importing.

---

## Key takeaways

- Make RStudio comfortable and readable.
- Learn a few shortcuts and use them daily.
- Export plots and tables deliberately.
- Always check imported data.

## Common pitfalls recap

- Forgetting to check the delimiter when importing.
- Using non‑monospace fonts that misalign code.
- Skipping basic checks after import.

## Mini‑project

Pick one dataset you have imported and:

1. Check it with `str()` and `summary()`.
2. Save a small summary table to CSV.
3. Export one plot and write a sentence describing it.
