## Exercise Solutions: Wrangling the Amniote Life History Database



In this exercise you used the "Amniote Life History Database" ^[https://esajournals.onlinelibrary.wiley.com/doi/10.1890/15-0846R.1] to investigate some questions about life history evolution.

The questions were:
(1) what are the records and typical life spans in different taxonomic classes? [*what is the longest, shortest and median life span in birds, mammals and reptiles?*]
(2) is there a positive relationship between body mass and life span? [*do big species live longer than small ones?*]; 
(3) is there a trade-off between reproductive effort and life span? [*do species that reproduce a lot have short lives, so there is a negative relationship between reproduction and life span?*]; 
(4) is this trade-off universal across all Classes? [*does the trade-off exist in birds, reptiles and amphibians?*]

The database is in a file called `Amniote_Database_Aug_2015.csv` in the course data folder. The missing values (which are normally coded as `NA` in R) are coded as "`-999`". The easiest way to take care of this is to specify this when we import the data using the `na.strings` argument of the `read.csv` function. Thus we can import the data like this:

```{r,eval=FALSE}
amniote <- read.csv("CourseData/Amniote_Database_Aug_2015.csv",na.strings = "-999")
```

Let's make a start...

1. When you have imported the data, use `dim` to check the dimensions of the whole data frame (you should see that there are `r ncol(amniote)` columns and `r nrow(amniote)` rows). Use `names` to look at the names of all columns in the data in `amniote`.

```{r}
dim(amniote)
names(amniote)
```

2. We are interested in longevity (lifespan) and body size and reproductive effort and how this might vary depending on the taxonomy (specifically, with Class). Use `select` to pick relevent columns of the dataset and discard the others. Call the new data frame `x`. 
The relevant columns are the taxonomic variables (`class`, `species`) and `longevity_y`, `litter_or_clutch_size_n`, `litters_or_clutches_per_y`, and `adult_body_mass_g`.


```{r}
x <- amniote %>%
  select(class,genus,species,
         longevity_y, adult_body_mass_g,
         litter_or_clutch_size_n,litters_or_clutches_per_y )
```

3. Take a look at the first few entries in the `species` column. You will see that it is only the *epithet*, the second part of the *Genus_species* name, that is given.  
Use `mutate` and `paste` to convert the `species` column to a *Genus_species* by pasting the data in `genus` and `species` together. To see how this works, try out the following command, `paste(1:3, 4:6)`. After you have created the new column, remove the `genus` column (using `select` and `-genus`).


```{r}
x <- x %>% 
  mutate(species = paste(genus,species)) %>%
  select(-genus)
head(x)
```

4. What is the longest living species in the record? Use `arrange` to sort the data from longest to shortest longevity (`longevity_y`), and then look at the top of the file using `head` to find out. (hint: you will need to use reverse sort (`-`)). Cut and paste the species name into Google to find out more!

```{r}
x <- x %>% arrange(-longevity_y)
head(x)
```

5. Do the same thing but this time find the shortest lived species.

```{r}
x <- x %>% arrange(longevity_y)
head(x)
```

6. Use `summarise` and `group_by` to make a table summarising `min`, `median` and `max` life spans (`longevity_y`) for the three taxonomic classes in the database. Remember that you need to tell R to remove the `NA` values using a `rm.na = TRUE` argument.

```{r}
x %>% 
  group_by(class) %>%
  summarise(min = min(longevity_y,na.rm = TRUE),
            median = median(longevity_y,na.rm = TRUE),
            max = max(longevity_y,na.rm = TRUE))
```

7. Body size is thought to be associated with life span. Let's treat that as a hypothesis and test it graphically. Sketch what would the graph would look like if the hypothesis were true, and if it was false. Plot `adult_body_mass_g` vs. `longevity_y` (using base R graphics). You should notice that this looks a bit messy.

```{r}
plot(x$adult_body_mass_g,x$longevity_y)
```

8. Use `mutate` to create a new `log`-transformed variables, `logMass` and `logLongevity`. Use these to make a "log-log" plot. You should see that makes the relationship more linear, and easier to "read".

```{r}
x <- x %>%
  mutate(logMass = log(adult_body_mass_g),
         logLongevity = log(longevity_y))

plot(x$logMass,x$logLongevity)
```

9. Is there a trade-off between reproductive effort and life span? Think about this as a hypothesis - sketch what would the graph would look like if that were true, and if it was false. Now use the data to test that hypothesis: Use `mutate` to create a variable called `logOffspring` which is the logarithm of number of litters/clutches per year multiplied by the number of babies in each litter/clutch . Then plot `logOffspring` vs. `logLongevity`.

```{r}
x <- x %>%
  mutate(logOffspring = log(litter_or_clutch_size_n * litters_or_clutches_per_y))

plot(x$logOffspring,x$logLongevity)
```

10. To answer the final question (differences between taxonomic classes) you could now use `filter` to subset to particular classes and repeat the plot to see whether the relationships holds universally. 

```{r}
aves <- x %>%
  filter(class == "Aves")

plot(aves$logOffspring,aves$logLongevity)
title("Aves")
```

```{r}
mammalia <- x %>%
  filter(class == "Mammalia")

plot(mammalia$logOffspring,mammalia$logLongevity)
title("Mammalia")
```

```{r}
reptilia <- x %>%
  filter(class == "Reptilia")

plot(reptilia$logOffspring,reptilia$logLongevity)
title("Reptilia")
```
