# Combining data sets

Handling data is often not limited to single data sets. One common task is to combine two (or more) data sets. For example, one data set might include observations from a field study, while another might have information about the weather during the study period or site-specific information. It is therefore useful to be able to combine these data sets to add information from the second table to the first.

**R** does this with `dplyr` join functions such as `left_join`. In the next section you will first learn how joins work by following an example that asks a research question that can only be answered by combining two data sets. After that, you will work on your own to do a similar analysis without explicit instructions (i.e. you will need to figure out how to apply the method to new data).

## Using `join`

By following this example you will learn how to combine two data sets to create a new data set for a conservation-related question: "**Does threat status vary with species' generation times?**" 

This question is crucial to conservation biologists because it helps us generalise our ideas about what drives extinction risk. In other words, if we can say "*species with slow life histories tend to be more threatened*", that gives useful information for planning. For example, imagine we have some species that have not yet been assessed (we don't know if they are threatened or not). Should we focus attention on the one with a short generation time, or the one with a long generation time?

To answer the question we will need to import two large data sets, tidy them up a bit, and then combine them for analysis.

Let's start with the "Amniote Life History Database" ^[See [https://esajournals.onlinelibrary.wiley.com/doi/10.1890/15-0846R.1](https://esajournals.onlinelibrary.wiley.com/doi/10.1890/15-0846R.1)], which is a good source of life history data. We have encountered this database before. Recall that the missing values (which are normally coded as `NA` in R) are coded as "`-999`". The easiest way to take care of this is to specify this when we import the data using the `na.strings` argument of the `read.csv` function. Thus we can import the data like this:


```{r 1-02-combiningDatasets-1}
amniote <- read.csv("CourseData/Amniote_Database_Aug_2015.csv",
  na.strings = "-999"
)
```

We can filter on the taxonomic `class` to subset to mammals. Then, to address our question, we want data on generation time for mammals. Generation time is often measured as the average age at which females reproduce, so we can get close to that with `female_maturity_d`. We will first `select` these columns, along with `genus` and `species`. We can combine these two taxonomic variables using `mutate` and `paste` to get our Latin binomial species name.

We have previously learned that log transforming such variables is useful, so we can use `mutate` again to do this transformation.

Finally, we can use `na.omit` to get rid of entries with missing values (which we cannot use). This is not essential, but keeps things more manageable.

```{r 1-02-combiningDatasets-2}
mammal <- amniote %>%
  filter(class == "Mammalia") %>%
  # get the mammals only
  select(genus, species, female_maturity_d) %>%
  # get useful columns
  mutate(species = paste(genus, species)) %>%
  select(-genus) %>%
  mutate(logMaturity = log(female_maturity_d)) %>%
  na.omit()
```


Let's take a quick look at what we have:

```{r 1-02-combiningDatasets-3}
head(mammal)
```

Looks good. Now let's import the IUCN Red List data.


```{r 1-02-combiningDatasets-4}
redlist <- read.csv("CourseData/MammalRedList.csv")
```

Let's take a look at that.

```{r 1-02-combiningDatasets-5}
names(redlist)
unique(redlist$Red.List.status)
```

There's a lot of information there but what we really need is the Latin binomial (for which we need `genus` and `species`) and the threat status `Red.List.status`. 

R treats categorical variables (`factor` variables) as alphabetical, but in this case the red list status has a meaning going from low threat (Least Concern - LC) to Critically Endangered (CR) and even Extinct in the Wild (EX) at the other end of the spectrum. We can define this ordering using `mutate` with the `factor` function. 

```{r 1-02-combiningDatasets-6}
redlist <- redlist %>%
  mutate(species = paste(Genus, Species)) %>%
  select(species, Red.List.status) %>%
  mutate(Red.List.status = factor(Red.List.status,
    levels = c("LC", "NT", "VU", "EN", "CR", "EW", "EX")
  ))

head(redlist)
```

Now we can combine this with the life history data from above using `left_join`. 

```{r 1-02-combiningDatasets-7}
x <- left_join(mammal, redlist, by = "species")
```

Let's take a look at what we have now:

```{r 1-02-combiningDatasets-8}
head(x)
summary(x)
```

You can see that there are `r sum(is.na(x$Red.List.status))` missing values for the Red List status. These are either species that have not yet been assessed, or cases where there are mismatches in the species names between the two databases. We will ignore this problem today. 

Before plotting, I will also use `filter` to remove species that are extinct (status = "EX" and "EW"). To do this I use the `%in%` argument to match a vector of values. Because I want to NOT match them I negate the match using `!`. 

```{r 1-02-combiningDatasets-9}
x <- x %>%
  filter(!Red.List.status %in% c("EX", "EW"))
```

Let's now plot the data to answer the question.

```{r 1-02-combiningDatasets-10}
plot(x$Red.List.status, x$logMaturity, ylab = "Maturity")
```

What can we see? If you focus on the median values, it looks like there is a weak positive relationship between this life history trait and threat status: animals with slower life histories tend to be more threatened.

## Using `pivot_longer`

Sometimes data are not arranged in a way that makes them easy to use in R. For example, data could be arranged so that the column headings are themselves data (e.g. treatments, sex of individuals).

To illustrate this I will use a data set on heights of men and women.

```{r 1-02-combiningDatasets-11}
heights <- read.csv("CourseData/heights.csv")
```
Let's take a look:

```{r 1-02-combiningDatasets-12}
heights
```

I would like to make a boxplot, but it is not possible to easily do it with the data arranged in this format. What I need to do is "unpack" or rearrange the data to add another column for sex. This will make the data frame twice as long, and less wide.

There is a convenient function called `pivot_longer` in the `tidyr` package that will do this for you. 
You can pipe data into the function, then tell it which columns you would like to move, and then give it the name of the new column that contains data that **was** in the column heading, and the name of the column containing the data. 

```{r 1-02-combiningDatasets-13}
newHeights <- heights %>%
  pivot_longer(
    cols = c(heightMale, heightFemale),
    names_to = c("Sex"), values_to = "Height"
  )

newHeights
```

We are nearly done. But this is not perfect because the names in the `Sex` column are not right. We can fix this in a couple ways. Here's one easy way using the function `gsub`. The `gsub` function ("general substitution") finds text and replaces it with other text. In this case we want to find "height" and replace it with nothing ("").

So we can now complete the job with a `mutate` command, and make sure it is recognised as a categorical variable (a `factor`), like this:

```{r 1-02-combiningDatasets-14}
newHeights <- heights %>%
  pivot_longer(
    cols = c(heightMale, heightFemale),
    names_to = c("Sex"), values_to = "Height"
  ) %>%
  mutate(Sex = gsub(
    pattern = "height", replacement = "",
    x = Sex
  )) %>%
  mutate(Sex = as.factor(Sex))

newHeights
```

Now we can plot those data more easily.

```{r 1-02-combiningDatasets-15}
plot(newHeights$Sex, newHeights$Height)
```

This data manipulation is useful surprisingly often.

## Exercise: Temperature effects on egg laying dates

```{r 1-02-combiningDatasets-16, eval = TRUE, echo=FALSE, warning=FALSE, message=FALSE}
# Here I am preparing the data for the course
sduBirds <- read.csv("CourseData/sduBirds.csv")

# Process to get first egg date
# First egg, data preparation ------------
xx <- sduBirds %>%
  mutate(nEggs = as.numeric(as.character(nEggs))) %>%
  select(Timestamp, Year, Day, boxNumber, nEggs) %>%
  filter(nEggs > 0) %>%
  group_by(Year, boxNumber) %>%
  filter(Day == min(Day)) %>%
  ungroup() %>%
  mutate(estimatedFirstEggDate = Day - nEggs, dummy = 1) %>%
  mutate(Year = as.factor(Year)) %>%
  arrange(estimatedFirstEggDate) %>%
  arrange(Year) %>%
  group_by(Year) %>%
  mutate(cumulativeCount = cumsum(dummy)) %>%
  select(-dummy) %>%
  ungroup() %>%
  mutate(Year = as.numeric(as.character(Year))) %>%
  select(boxNumber, Year, estimatedFirstEggDate) %>%
  rename(firstEgg = estimatedFirstEggDate) %>%
  arrange(boxNumber, Year) %>%
  spread(Year, firstEgg)

names(xx)[2:ncol(xx)] <- paste0("y", names(xx)[2:ncol(xx)])
write.csv(xx, "CourseData/eggDates.csv", row.names = FALSE)

# Prep weather
weather <- read.csv("DataSetLibrary/SDUweather.csv") %>%
  select(YEAR, MONTH, DAY, TEMP)

write.csv(weather, "CourseData/AarslevTemperature.csv", row.names = FALSE)
```

Data have been collected on great tits (musvit) at SDU for several years. Your task today is to analyse these data to answer the question: *is egg laying date associated with spring temperature?* The idea here is that warmer springs can lead to earlier egg laying, which could have negative consequences for the population if their caterpillar food source does not keep pace with the change.

You are provided with two data sets: one on the birds and another on weather. You will need to process these using tools in the `dplyr` package, and combine them (using `left_join`) for analysis.

The first data set, `eggDates.csv`, is data from the SDU birds project. The data are arranged in columns where each column is a year and each row is a nest. The data in each column is the day of the year that the first egg in the nest was laid.

These data do NOT fulfil the "tidy data" standard where each variable gets a column. In this case, a single variable (first egg date) gets many columns (one for each year), and column headers are data (the years). The data will need to be processed before you can analyse it.

You will need to use `pivot_longer` to fix this issue so that you produce a version of the data with three columns: `boxNumber`, `Year`, and `dayNumber` (or similar names).

The second dataset, `AarslevTemperature.csv`, is a weather dataset from Årslev near Odense. This dataset includes daily temperature records for several years. You will need to `summarise` this data to obtain a small dataset that has the weather of interest - average temperature in the months of February to April for each year.

To answer the question, you will need to join these data sets together.

1. Import the data and take a look at it  with `head` or `str`.

2. Use `pivot_longer` to reformat the data. This might take a bit of trial and error - don't give up! 

Maybe this will help: The first argument in the `pivot_longer` command (`cols`) tells R which columns contain the data you are interested in (in this case, these are `y2013`,`y2014` etc). Then the `names_to` argument tells R what you want to name the new column from this data (in this case, `Year`). Then, the `values_to` argument tells R what the data column should be called (e.g. `Day`). In addition, there is a useful argument called `names_prefix` that will remove the part of the column name  (e.g. the `y` of `y2013`)

You should also make sure that the `Year` column is recognised as being a numeric variable rather than a character string. You can do this by adding a command using `mutate` and `as.numeric`, like this `mutate(Year = as.numeric(Year))`

You should end up with a dataset with three columns as described above.


3. Calculate the mean egg date per year using `summarise` (remember to `group_by` the year first). Take a look at the data.

4. Import the weather data and take a look at it with `head` or `str`.

5. Use `filter` to subset to the months of interest (February-April) and then `summarise` the data to calculate the mean temperature in this period (remember to `group_by` year). Look at the data. You should end up with a dataset with two columns - `year` and `meanSpringTemp`.

6. Join the two datasets together using `left_join`. You should now have a dataset with columns `boxNumber`, `Year`, `dayNumber`, and `meanSpringTemp` (or equivalent names).

7. Plot a graph of `meanSpringTemp` on the x-axis and `dayNumber` on the y-axis.

Now you should be able to answer the question we started with: is laying date associated with spring temperatures.
