# Visualising data with `ggplot`


In this chapter you will be guided through using the `ggplot2` package to make some pretty plots. You will therefore need the `ggplot2` package to make this work. Remember, you can load packages like this:


```{r,message=FALSE}
library(ggplot2)
```

We will use the SDU birds clutch size data that we produced at the end of the "[Data wrangling with dplyr]" chapter for these examples. You can find the data set via the Course Data Dropbox link.


```{block, type="do-something"}
Remember to set your working directory, and start a new script. I am assuming that you have saved your data in a folder called "CourseData" inside your working directory. 
```

```{r}
df <- read.csv("CourseData/SDUClutchSize.csv")
```

## Histograms

The `ggplot` function expects two main arguments (1) the data and (2) the **aesthetics**. The aesthetics are the variables you want to plot, and associated characteristics like colours, groupings etc. The first argument is for the data, then the aesthetics are specified within the `aes(...)` argument. These usually include an argument for `x` which is normally the variable that appears on the horizontal axis, and (often) `y` which is usually the variable on the vertical axis. The details of this depend on the type of plot you are making.

After setting up the plot the graphics are added as **geometric layers** or **geoms**. There are many of these available including `geom_histogram`, `geom_line`, `geom_point` etc.

I will illustrate the construction of a simple plot by making a histogram of the clutch size of all the nests in the dataset.

```{r}
ggplot(df, aes(x = clutchSize))
```

This produces an empty plot because we have not yet specified what kind of plot we want. We want a histogram, so we can add this as follows. I have set `binwidth` to be 1 because we know we are dealing with counts between just 1 and 14. Try altering the `binwidth`.

```{r}
ggplot(df, aes(x = clutchSize)) +
  geom_histogram(binwidth = 1)
```
We know that we have two species here and we would like to compare them. This is done within the aesthetic argument. The default is that the bars for different categories are stacked on top of each other. This is good in some cases, but probably not here.

```{r}
ggplot(df, aes(x = clutchSize, fill = species)) +
  geom_histogram(binwidth = 1, position = "dodge")
```

You can immediately see that there are far fewer blue tit nests than great tit ones. But you can also see that the center of mass for blue tits is further to the right than great tits.

To make it easier to compare distributions with very different counts, we can put density on the y-axis instead of the default count using the argument `stat(density)`.


```{r}
ggplot(df, aes(x = clutchSize, fill = species, stat(density))) +
  geom_histogram(binwidth = 1, position = "dodge")
```

An alternative approach would be to overlay the two sets of bars (using `position = "identity"`) and set the colours to be slightly transparent (using `alpha = 0.7`) so that you can see the overlapping region clearly. 

```{r}
ggplot(df, aes(x = clutchSize, fill = species, stat(density))) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.7)
```

It is very clear from this plot that blue tits tend to have bigger clutch sizes than great tits. Is this difference *statistically significant*? We will look at testing this in a future class - for now we will be satisfied with our visualisation.

## "Facets" - splitting data across panels

You should recall that there were several years of data represented here.
`ggplot` has a very clever way of splitting up the plot to examine this.

```{r}
ggplot(df, aes(x = clutchSize, fill = species, stat(density))) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.7) +
  facet_grid(. ~ Year)
```

You could split the data up by species in a similar way, as yet another way of visualising the difference between species:

```{r}
ggplot(df, aes(x = clutchSize)) +
  geom_histogram(binwidth = 1) +
  facet_grid(species ~ .)
```

```{block, type="do-something"}
You can change whether the separate graphs are presented in a rows or columns by changing the order of the argument: ` facet_grid(species~.)` or ` facet_grid(.~species)`. Try it.
```



## Box plots

Box plots are suitable for cases where one variable is categorical with 2+ levels, and the other is continuous. Therefore, another way to look at these distributions is to use a box plot.

In a box plot the box shows the quartiles (i.e. the 25% and 75% quantiles) within which 50% of the data are found. The horizontal line in the box is the *median*, Then the whiskers extend from the smallest to largest value *unless they are further than 1.5 times the interquartile range (the length of the box)* away from the edge of the box, in which case they are individually shown as outlier points.

To plot them using `ggplot` you must use a `geom_boxplot` layer. 
The categorical variable is normally placed on the x-axis so is placed as `x` in the `aes` argument, while the continuous variable is on the `y` axis.

```{r}
ggplot(df, aes(x = species, y = clutchSize)) +
  geom_boxplot()
```


Some researchers argue that it is a good idea to add the data as points to these plots as "full disclosure" of what the underlying data look like. These can be added with a `geom_jitter` layer (jitter is random noise added in this case to the horizontal axis). You should set `width` and `alpha` arguments to make it look nice.

```{r,echo = -1,cache = TRUE}
set.seed(42)
ggplot(df, aes(x = species, y = clutchSize)) +
  geom_boxplot() +
  geom_jitter(width = .2, alpha = 0.5, colour = "black", fill = "black")
```

```{block, type="do-something"}
Try splitting the data into different years using `facet_grid` with the box plot. 
```


## Lines and points

Perhaps not surprisingly lines and points can be added with the geoms, `geom_line` and `geom_point` respectively.
To illustrate this we will make a plot showing how clutch size changes among years.
First we will use `summarise` to create a dataset with the mean clutch size. We'll start simply, by looking at only great tits.


```{r}
GTclutch <- df %>%
  filter(species == "GT") %>%
  group_by(Year) %>%
  summarise(meanClutchSize = mean(clutchSize))
```

Then you can plot this like this.

```{r}
ggplot(GTclutch, aes(x = Year, y = meanClutchSize)) +
  geom_line()
```

I think this looks OK, but we should add both species. I'll first need to produce a mean clutch size dataset that includes both species.


```{r}
meanClutch <- df %>%
  group_by(species, Year) %>%
  summarise(meanClutchSize = mean(clutchSize))
```

Now I can do the plot again. The only difference to the command is that I need to tell R that I want to colour the lines by species (`colour = species`).

```{r}
ggplot(meanClutch, aes(x = Year, y = meanClutchSize, colour = species)) +
  geom_line()
```

I can improve on this by (1) changing the y axis limits (using `ylim`) so that it goes through the full range of my data (0 - 14); (2) adding points (using a `geom_point` layer) where my actual data values are; (3) adding a nicely formatted axis label (using `ylab`); adding a title (`ggtitle`)

```{r}
ggplot(meanClutch, aes(x = Year, y = meanClutchSize, colour = species)) +
  geom_line() +
  geom_point() +
  ylim(0, 14) +
  ylab("Mean clutch size") +
  ggtitle("Clutch size data from SDU Campus")
```

## Scatter plots

Finally, lets make a scatter plot.
The SDU bird data are not suitable for this type of plot so we'll use the data from a few days ago on suburban bird diversity.

```{r}
df <- read.csv("CourseData/suburbanBirds.csv")
```

Take a look at the data to remind ourselves what it looks like

```{r}
head(df)
```

These data show the result of standardised bird surveys at housing developments of different ages in California. The surveys were carried out in 1975, and the data includes the `Year` and number of individual birds seen `nIndividuals` and number of species seen `nSpecies`. The question being addressed is "How does the age of the housing development affect the number of species?"

To investigate this we should first add a new variable for `Age` to the data set. We can do this using the `mutate` function from `dplyr`. This function creates new variables, for example by manipulating existing ones.

```{r}
df <- mutate(df, Age = 1975 - Year)
```

When we have created this variable we can plot the data. For aesthetic reasons I also would like to set the limits on the y-axis to go extend to zero, and I would like to include proper labels on the axes.

```{r}
ggplot(df, aes(x = Age, y = nSpecies)) +
  geom_point() +
  ylim(0, 15) +
  xlab("Age of development") +
  ylab("Bird species richness")
```

This shows very clearly that older developments have more species, but it also appears to show that there is an asymptote around 13 species.

Compare this plot to the one you made with base graphics in a previous class.
