# Visualising data with `ggplot2`

**Learning goals**

By the end of this chapter you should be able to:

- Build plots with `ggplot()` and geoms.
- Compare groups using histograms and box plots.
- Use facets to split plots into panels.
- Plot trends over time with lines and points.
- Decide when a bar plot is (and is not) appropriate.

**Prerequisites**

- You should be comfortable with importing data and using `filter()`, `select()`, `mutate()`, and `summarise()`.

**A tiny motivating example**

Let’s start with a simple scatter plot. Run this and see what happens:

```{r 2-01-visualisingData-1, message=FALSE}
library(tidyverse)

mydata <- data.frame(
  body_length_cm = c(4.2, 5.1, 5.6, 6.8, 7.4),
  body_mass_g = c(9.5, 12.1, 14.0, 18.2, 21.5)
)

ggplot(mydata, aes(x = body_length_cm, y = body_mass_g)) +
  geom_point()
```

We have already made a plot: data + aesthetics + geom. That pattern will repeat throughout the chapter.

---

## The data we will use

We will use the SDU clutch size data created at the end of the **Data wrangling** chapter. The file should be in `CourseData/SDUClutchSize.csv`.

```{r 2-01-visualisingData-2}
clutch <- readr::read_csv("CourseData/SDUClutchSize.csv")
```

**Checkpoint:** You should see columns like `species`, `Year`, and `clutchSize`.

---

## Histograms

**What we’re about to do:** We will look at the distribution of clutch sizes. Histograms are the simplest way to see how values are spread.

```{r 2-01-visualisingData-3}
clutch %>%
  ggplot(aes(x = clutchSize)) +
  geom_histogram(binwidth = 1)
```

**What just happened:** Each bar shows how many nests have a particular clutch size (or range of sizes).

- **What to look for:**
  - Where is the distribution centred?
  - Is it roughly symmetric or skewed?
  - Are there values that are rare or extreme?

**Try this:** Change `binwidth` to 0.5 or 2. What changes in the shape?

**Common mistake:** If your plot is blank, check that `clutchSize` exists with `names(clutch)`.

**Takeaway:** Histograms show the shape of a distribution quickly.

---

## Comparing groups in a histogram

**What we’re about to do:** We will compare clutch sizes between species.

```{r 2-01-visualisingData-4}
clutch %>%
  ggplot(aes(x = clutchSize, fill = species)) +
  geom_histogram(binwidth = 1, position = "dodge")
```

**What just happened:** The bars are separated by species, so you can compare their shapes.

- **What to look for:**
  - Do the peaks occur at different clutch sizes?
  - Is one species more common (taller bars overall)?

**Caution:** Counts depend on sample size. If one species has more nests, it will look bigger even if the *shape* is similar.

**Try this:** Switch to densities so the shapes are comparable:

```{r 2-01-visualisingData-5}
clutch %>%
  ggplot(aes(x = clutchSize, fill = species, y = after_stat(density))) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.7)
```

**Takeaway:** Use densities when sample sizes differ.

---

## Facets (split across panels)

**What we’re about to do:** We will split plots by year so each year gets its own panel.

```{r 2-01-visualisingData-6}
clutch %>%
  ggplot(aes(x = clutchSize, fill = species, y = after_stat(density))) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.7) +
  facet_grid(. ~ Year)
```

**What just happened:** Each year is shown in a separate panel. This makes year-to-year differences easier to see.

- **What to look for:**
  - Does the distribution shift over time?
  - Are some years more variable than others?

**Try this:** Flip the facet so species are rows:

```{r 2-01-visualisingData-7}
clutch %>%
  ggplot(aes(x = clutchSize)) +
  geom_histogram(binwidth = 1) +
  facet_grid(species ~ .)
```

**Takeaway:** Facets let you compare groups without overplotting.

---

## Box plots

**What we’re about to do:** We will compare clutch size between species using box plots.

```{r 2-01-visualisingData-8}
clutch %>%
  ggplot(aes(x = species, y = clutchSize)) +
  geom_boxplot()
```

**What just happened:** The box shows the middle 50% of values; the line is the median.

- **What to look for:**
  - Are medians different between species?
  - Are spreads similar or very different?

**Try this:** Add jittered points for transparency.

```{r 2-01-visualisingData-9, echo = FALSE}
set.seed(42)
clutch %>%
  ggplot(aes(x = species, y = clutchSize)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5)
```

**Common mistake:** If your points overlap too much, increase `width` or reduce `alpha`.

**Takeaway:** Box plots are usually better than bar plots for comparing distributions.

---

## Lines and points (trends over time)

**What we’re about to do:** We will summarise clutch size by year and plot a time trend.

```{r 2-01-visualisingData-10}
mean_clutch <- clutch %>%
  group_by(species, Year) %>%
  summarise(mean_clutch_size = mean(clutchSize), .groups = "drop")
```

```{r 2-01-visualisingData-11}
mean_clutch %>%
  ggplot(aes(x = Year, y = mean_clutch_size, colour = species)) +
  geom_line() +
  geom_point() +
  ylim(0, 14) +
  ylab("Mean clutch size") +
  ggtitle("Clutch size data from SDU campus")
```

**What just happened:** Each point is a year’s mean clutch size; lines connect years.

- **What to look for:**
  - Do the two species move together or differently?
  - Are there years that stand out?

**Caution:** A mean hides variability. For deeper analysis, check sample sizes or plot distributions per year.

**Takeaway:** Lines are good for ordered values like time.

---

## Scatter plots

**What we’re about to do:** We will plot how bird species richness changes with suburb age.

```{r 2-01-visualisingData-12}
birds <- readr::read_csv("CourseData/suburbanBirds.csv") %>%
  mutate(Age = 1975 - Year)
```

```{r 2-01-visualisingData-13}
birds %>%
  ggplot(aes(x = Age, y = nSpecies)) +
  geom_point() +
  ylim(0, 15) +
  xlab("Age of development") +
  ylab("Bird species richness")
```

**What just happened:** Each point is a suburb. The plot shows older suburbs tend to have more species.

- **What to look for:**
  - Is the relationship roughly linear or does it level off?
  - Are there outliers that might drive the pattern?

**Try this:** Add a smooth trend line with `geom_smooth()` and see if the relationship looks curved.

**Takeaway:** Scatter plots are the starting point for relationship questions.

---

## Bar plots (use with care)

**What we’re about to do:** We will see why bar plots can be misleading and when they are acceptable.

Bar charts can hide the shape of data. They are fine for **counts**, but risky for **means**.

```{r 2-01-visualisingData-14, echo = FALSE, fig.width = 6, fig.height = 2, fig.align = "center", fig.cap = "Problems with bar plots."}
library(patchwork)
set.seed(42)
n <- 50
mv <- 2
v <- 3
shape <- (mv^2) / v
scale <- v / mv

df1 <- data.frame(
  x = rep(c("G1", "G2"), each = n),
  y = c(rnorm(n, mean = mv), rgamma(n, shape = shape, scale = scale))
)

df2 <- df1 %>%
  group_by(x) %>%
  summarise(meanVal = mean(y), sdVal = sd(y), .groups = "drop")

A <- ggplot(df2, aes(x = x, y = meanVal)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_errorbar(
    aes(ymin = meanVal - sdVal / sqrt(n), ymax = meanVal + sdVal / sqrt(n)),
    width = 0.2
  ) +
  theme_minimal() +
  ggtitle("A") +
  ylab("y")

B <- ggplot(df1, aes(x = x, y = y)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(size = 2, width = 0.2) +
  theme_minimal() +
  ggtitle("B") +
  ylab("y")

A | B
```

**Interpretation:** The bar plot hides the skew and outliers visible in the box plot.

- **What to look for:**
  - How much information disappears when you only show the mean?

**When bar plots are OK:** Counts are a good example.

```{r 2-01-visualisingData-15}
clutch %>%
  count(species, name = "sample_size") %>%
  ggplot(aes(x = species, y = sample_size)) +
  geom_bar(stat = "identity") +
  labs(title = "Sample sizes by species", x = "Species", y = "Sample size")
```

**Try this:** Plot counts by year and compare a bar plot to a line plot. Which tells the story more clearly?

**Takeaway:** Use bar plots for counts, not for means.

---

## Key takeaways

- `ggplot(data, aes(...)) + geom_*()` is the core pattern.
- Use histograms and box plots to understand distributions.
- Use facets to compare groups without clutter.
- Use lines and points for ordered data like time.
- Use bar plots mainly for counts.

## Common pitfalls recap

- Mapping the wrong variable to `x` or `y` (check your `aes()` call).
- Forgetting `geom_*()` (you’ll get an empty plot).
- Using bar plots for means without showing the distribution.

## Mini-project

Choose one dataset from `CourseData` and produce three plots:

1. A distribution plot (histogram or box plot).
2. A relationship plot (scatter or line).
3. A grouped comparison (facets or colour).

Write 2–3 sentences under each plot describing what it shows and one thing you would check next.
