# Exam 2022



```{block, type="do-something"}

**Read these instructions carefully**

This exam includes four questions that test different aspects of your learning during this course: data wrangling, data visualisation and statistics. Each question is broken down into a number sub-questions. 

Your work should be handed in as a single PDF. 
The number of each question should be clearly indicated, and the answer to each question should start on a new page.

For each question you must provide the R code you used to answer the question. The code should include comments to explain what you are doing. The code should be provided as text using a fixed-width font such as `Courier`. The rest of your answers should be in another font (e.g. Times New Roman, Cambria). Please use *text* rather than a screenshot of your code.

* Plots and tables should have appropriate captions. 
* Plots should be produced using `ggplot`.
* Remember that you can make "panels" of plots (e.g. Fig 1A, B)
* Axis labels are important. Sometimes you may want to edit them to be different from a data column name.
* Reporting of any statistics should be appropriate to the type of analysis you have done. There are examples in the course materials.
* Reporting of methods and results should be written in the style of a scientific paper (again, there are examples in the course materials).

*If you don't understand any aspects of the questions, please ask for help!*

**Hand in deadline is 9th January 2023 at 23:59 CET**

**You MUST submit your work via itsLearning (not email!)**
```

```{r,echo=FALSE,message=FALSE}
library(tidyverse)
library(ggfortify)

options(dplyr.summarise.inform = FALSE)
```

*****

**1) Kaunas weather (10 points)**

The dataset `kaunasWeather.csv` is weather data obtained from near Kaunas in Lithuania, starting in 1953 and ending in 2022.

Each row of the dataset represents a single day.

`YEAR`, `MONTH`, `DAY` give the calendar dates.
`YDAY` is the day of the year.
`TEMPMEAN` is the mean temperature of the day (Celsius).
`TEMPMIN` is the minimum temperature of the day (Celsius).
`TEMPMAX` is the minimum temperature of the day (Celsius).
`PRCP` is the precipitation (rain and melted snow) (millimeters).
`MXSPD` is the maximum sustained wind speed (meters per second).

a) Produce a graph with points and lines showing how the monthly mean value of the daily temperature varies through the year (i.e. 12 points, 1 for each month).

Import the data:
```{r}
kaunas <- read.csv("CourseData/kaunasWeather.csv")
```

Calculate the montly mean of the daily temperature. I need 12 points - one for each month.

```{r}
library(tidyverse)
kaunas_summary <- kaunas %>%
  group_by(MONTH) %>%
  summarise(meanTemp = mean(TEMPMEAN))
```

Make the plot:

```{r}
ggplot(kaunas_summary, aes(x = MONTH, y = meanTemp)) +
  geom_point() +
  geom_line() +
  xlab("Month") +
  ylab("Temperature (Celsius)") +
  scale_x_continuous(breaks = 1:12) +
  theme_minimal()
```
The figure caption could be: "Figure 1. The monthly mean temperature in Kaunas across all available data. The average temperatures are below zero in January and February, rising to a high of around 17.5C in July."


b) Produce a graph showing how mean temperature in spring (defined as March, April and May) have changed between 1960 and 2020 (i.e. one point per year).

First I need to filter the data to (A) only include spring months, and (B) only include years 1960-2020. Then I need to group it by YEAR, and then summarise to get the mean temperature.

```{r}
kaunas_springTemp <- kaunas %>% 
  filter(MONTH %in% 3:5) %>% 
  filter(YEAR %in% 1960:2020) %>% 
  group_by(YEAR) %>% 
  summarise(meanTemp = mean(TEMPMEAN))
```

```{r}
ggplot(kaunas_springTemp,aes(x = YEAR, y = meanTemp)) + 
  geom_point() +
  xlab("Year") + 
  ylab("Mean Spring Temperature (Celsius)")

```

The figure caption could be: "Figure 2. Mean spring temperature between 1960 and 2020 in Kaunas. Although there is a large amount of variation there is an increasing temperature trend over the 60 years of the data."

c) Produce a graph with box plots and jittered points showing daily temperature variation within each month in 2021. (there should be 12 boxes, one for each month).

For this I need to filter the data to include only 2021. Then I can make my plot:

```{r}
kaunas_2021 <- kaunas %>% 
  filter(YEAR == 2021) %>% 
  mutate(MONTH = as.factor(MONTH))

ggplot(kaunas_2021, aes(x = MONTH, y = TEMPMEAN))+
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(width = 0.1, colour = "grey30", alpha = 0.5) +
  xlab("Month")+ylab("Temperature (Celsius)")
```

The figure caption could be: "Figure 3. Daily temperature in 2021 in Kaunas. The individual grey points indicate the daily mean temperature, while the box plots summarise the distribution of daily values by month."

d) Produce a graph showing, with points, the number of days per year with storm speed winds (>14 m/s). (i.e. one point per year).

For this I need to count the number of times a wind speed value of >14 occurs per month. There is more than one way of doing this, but this is how I would do it:

```{r}
kaunas_storm <- kaunas %>% 
  filter(MXSPD > 14) %>% 
  group_by(YEAR) %>% 
  summarise(nDays = n())
```

Now I can plot that data:

```{r}
ggplot(kaunas_storm,aes(x = YEAR, y = nDays)) + 
  geom_point() +
  xlab("Year")+ 
  ylab("Number of days with wind speeds > 14m/s")
```

The figure caption could be: "Figure 4. Number of days per year where wind speeds exceed 14 m/s show a declining trend with time."



******

**2) Character dispacement in lizards (10 points)**


Character displacement is the phenomenon where differences among similar species with overlapping geographic distributions are accentuated in regions where the species co-occur. This means that a competitor's presence can influence a species's morphology (e.g., adult body size).

Researchers collected data (`Niveoscincus.csv`) to examine this phenomenon in an alpine lizard _Niveoscincus microlepidotus_ in Tasmania. The species' geographic range overlaps in some areas with a close relative _N. greeni_. The researchers measured the size (`snoutVentLength`) of _N. microlepidotus_ in two places: firstly, an area where the competitor species does not occur (`AllopatricZone`), and secondly, in an area where both species co-occur and actively compete (`OverlapZone`).

Use a randomisation test to test whether there is a significant difference in the mean size of _N. microlepidotus_ in the two areas.


a) plot the data (e.g., with a box plot)

```{r}
lizard <- read.csv("CourseData/Niveoscincus.csv")

ggplot(lizard, aes(x = factor(area), y = snoutVentLength)) +    geom_boxplot() +
  geom_jitter(width = .2, alpha = 0.4, colour = "black", fill = "black") +

  xlab("Area") +
  ylab("Length (mm)") + theme_minimal()
```


b) carry out a randomisation test to determine if there is a significant difference in mean size. Write (i) a brief method description and (ii) a summary of the results. 


First get the observed difference.

```{r}
observedDiff <- lizard %>%
  group_by(area) %>% 
  summarise(mean = mean(snoutVentLength)) %>% 
  pull(mean) %>%
  diff() 
```

The observed difference is `r observedDiff`

Now get the null distribution:

```{r}
shuffledDataframe <- data.frame(rep = 1:1000) %>%
  mutate(shuffledDiffs = replicate(
    1000,
    lizard %>%
      mutate(area = sample(area)) %>%
      group_by(area) %>%
      summarise(mean = mean(snoutVentLength)) %>%
      pull(mean) %>%
diff() ))
```


Now we can plot the observed value onto this null distribution and check the result.

```{r}
ggplot(shuffledDataframe, aes(x = shuffledDiffs)) +
  geom_histogram() +
  theme_minimal() +
  xlab("Difference between means") +
  geom_vline(xintercept = observedDiff)

```

The p-value is the number of times that the 

```{r}
(result <- table(shuffledDataframe$shuffledDiffs <= observedDiff))
```

The p-value is defined as the proportion of times that the values inthe null distribution is more extreme than the observed value. i.e. `r as.vector(result[2])`/`as.vector(sum(result))` = `r as.vector(result[2]/sum(result))`.



******


**3) Power in a new study (10 points)**

A recent study of 37 people in Copenhagen found that people produce a mean of 245.4 grams (sem = 10.1) of avoidable food waste per capita per week.

In connection with the UN Sustainable Development Goals you plan a new project based here in Odense that aims to reduce avoidable food waste by 25% through a combination of education and coaching. Odense municipality may be able to fund the project, but they need details on costs, which you estimate to be 125kr per participant. Therefore, you need to do a power analysis to determine the number of participants required to demonstrate a statistically significant reduction in food waste.

You design a study with two treatment groups: control and coached. The coached group will receive information and coaching, while the control group will receive no special training.

a) Conduct a power analysis based on the findings of the Copenhagen study to estimate the number of participants required to carry out your experiment with 80% power. 

b) Describe the methods of this power analysis.

c) Write up the power analysis results to justify your new project's costs.

******



**4) Weapon size and body size in crabs (20 points)**

Individuals compete to obtain resources where these resources, such as food or mates, are limited. Many factors, including morphological characteristics such as body size, determine an individual's resource-holding potential (RHP). During aggressive encounters, individuals compare their own RHP with their opponent before deciding whether to escalate their encounter: When an individual competes with a larger opponent, it will often concede the resources to its opponent without fighting. Prolonged fights are rare and occur only between evenly-matched individuals. 

Many studies have shown that body size or weight is a good predictor of the outcome of aggressive encounters, with larger individuals tending to win (often without a fight). In addition to body size, the evolution of weapons such as horns and antlers provides another method for individuals to signal their fighting prowess. The fact that there are two "indicators" of RHP (body size and weapon size) has generated debate about which is more important. For example, could a smaller animal with large horns beat a larger animal with smaller horns?

Researchers tested this idea using shore crabs (_Carcinus maenas_). First, they collected males of different sizes and measured their body size (carapace width) and weapon size (claw length). The researchers then staged fights between pairs of crabs (i.e., a "focal crab" and an "opponent crab") and observed the outcome. The winner was the crab that successfully climbed on top of its opponent or elicited repeated retreats from the other crab (the loser).

These data are stored in the file `crabFights.csv`. The data set has seven columns. The first two columns are the ID numbers of the focal and opponent crabs. Columns 3 and 4 give the carapace width (`focal_carapaceWidth`) and chela length (`focal_chelaLength`) of the focal crab in millimeters. Columns 5 and 6 provide the same information for the opponent crab. The final column (`fightOutcome`), gives the fight's outcome: `won` indicates that the focal crab won the fight, `lost` means that it lost the fight.


You aim to determine how the relative size (i.e., focal crab size divided by opponent crab size) of the body and the weapon of the crab influence the outcome of these fights.

a) Plot the data to show the relationship between carapace width and chela length.

b) Use a suitable statistical model to describe the relationship between carapace width and chela length. Add the model fit line to the plot produced in (a)

c) Calculate relative sizes (i.e., focal/opponent size) and then fit a suitable model (or models) to describe the relationship between size and fight outcome.

d) Describe the results of (c). What is the relative importance of the two size measures?


******
