# Exam 2021


```{block, type="do-something"}

This exam includes four questions that test different aspects of your learning during this course: data wrangling, data visualisation and statistics. Each question is broken down into a number sub-questions. 

Your work should be handed in as a single PDF. 

The number of each question should be clearly indicated, and the answers to each question should start on a new page.

For each question you must provide the R code you used to answer the question. The code should include comments to explain what you are doing. The code should be provided as text using a fixed-width font such as `Courier`. The rest of your answers should be in another font (e.g. Times New Roman, Cambria). Please use *text* rather than a screenshot of your code.

You can use the Microsoft Word template provided alongside these questions ([here](https://www.dropbox.com/s/0ecb2c4fl6ygui9/BB852_Exam_template.docx?dl=0)) as guidance.

* Plots and tables should have appropriate captions. 
* Plots should be produced using `ggplot`.
* Remember that you can make "panels" of plots (e.g. Fig 1A, B)
* Axis labels are important. Sometimes you may want to edit them to be different from a data column name.
* Reporting of any statistics should be appropriate to the type of analysis you have done. There are examples in the course materials.
* Reporting of methods and results should be written in the style of a scientific paper (again, there are examples in the course materials).

*If you don't understand any aspects of the questions, please ask for help!*

**Hand in deadline is 7th January 2022 at 12:00 CET (noon)**

**You MUST submit your work via itsLearning (not email!) Instructions for this will follow.**
```

```{r,echo=FALSE,message=FALSE}
library(tidyverse)
library(ggfortify)

options(dplyr.summarise.inform = FALSE)
```

*****

## 1) Moths and butterflies (10 points)


The dataset `lepi.csv` is a subset of Lepidoptera (moths and butterflies) species occurrence data for Denmark held by the Global Biodiversity Information Facility (GBIF) (https://www.gbif.org). It includes columns for taxonomy, geographic location, and time. The column called `dayOfYear` is the numeric day of the year (i.e. January 1st is day 1).

a) Identify the five most commonest species in the dataset and filter the data to just these species. 

Next, make a two panel plot with two graphs side-by-side. For plot A, make a graph with points joined by lines, that shows the yearly number of observations. The x-axis should be the year and the y-axis should be the number of observations (for all 5 species combined).

For plot B, make a single graph showing the number of observations per year for each of these species. Year should be on the horizontal axis, and number of observations should be on the vertical axis. Each species should be represented by a line with points (i.e. there should be five lines, each with a different colour).

b) Produce a table showing the minimum, maximum, mean and standard deviation of the number of observations per year for each species. i.e. column 1 should be species, column 2-5 should be min, max, mean and standard deviation. Arrange the data in descending order of mean number of observations (i.e. largest number first).



## 2) Home range size (10 points)

The home range size of great tits was assessed in two areas of woodland has been recorded to range between 0.06 to 14 hectares. This large variation may be influenced by the environment. To test this idea, researchers estimated the range size of birds from two different environments (1) mature broadleaf woodland, which is thought to be the best habitat for these birds and (2) pine plantation forest, which is thought to be inferior habitat. The hypothesis is that birds in poorer habitat need to have bigger home ranges in order to survive.

Test this hypothesis using the dataset `birdRangeSize.csv`

```{r}
set.seed(123)
birdRangeSize <- data.frame(birdID = 1:30,
                            habitat = rep(c("Broadleaf", "Pine"),each = 15)) %>% 
  mutate(rangeSize = round(c(rnorm(15,4,2),rnorm(15,7,3)),3))
  birdRangeSize
```




a) plot the data using a boxplot and overlaid jittered points.

```{r}
ggplot(birdRangeSize,aes(x = habitat, y = rangeSize)) + geom_boxplot() + geom_jitter()
```


b) carry out a randomisation test to determine if there is a significant difference in the range size in the two habitats. Write (i) a brief method description and (ii) a summary of the results. 

```{r}

```


##  3) Power in a field experiment (10 points)


Running speed is a key attribute of escape behaviour.  Golden-mantled ground squirrels (*Spermophilus lateralis*) and least chipmunks (*Tamias minimus*) live together on the mountains of North America. It has been hypothesised that the ground squirrels can utilise more habitat area than least chipmunks they are faster, and can therefore use areas that are more exposed to predation (they would be able to run away quicker). 

Your colleague studies one of these species, ground squirrels, and has measured running speed for a small number of individuals in this species only. She has sent you the data as a file called `runningSquirrels.csv`. 

You would like to test the hypothesis that the least chipmunks are slower than ground squirrels. You decide that a speed difference of 20% would be biologically meaningful. However, due to current travel restrictions you cannot go to the field site in Colorado, USA yourself. Instead, you must instruct a research assistant at the site to measure running speed in the chipmunks. They need to know how many of each species to measure.

Use the pilot study data to conduct a power analysis to help plan your study.


```{r}
set.seed(23)
ss <- 8
runningSquirrels <- data.frame(tagNumber = sample(100:500, size = ss)) %>% 
  mutate(speed = round(rnorm(ss,3.14,1),3))
```






a) Summarise the pilot study data to obtain mean and standard deviation.

b) Conduct a power analysis based on the pilot study data to estimate the number of samples required to carry out your experiment with 80% power. Describe the results of this power analysis.



**4) Biodiversity (20 points)**

There is a well-known relationship between habitat area and biodiversity - the "species-area relationship". In a nutshell, the number of species tends to increase as the area available increases. This relationship is clearly seen, for example, if we look at island biodiversity: larger islands support more species then small ones. One potential mechanism for this observed pattern is that larger islands tend to have more variety of different habitats, and therefore more niches available for species to occupy (more niches = more species).

The dataset `roundabouts.csv` shows the results of a study carried out in an urban environment to investigate these ideas.  During the study, the number of beetle species (`nSpecies`) living on roundabouts and other "islands" of vegetation of different sizes (`area` in $m^2$) in a sea of concrete and tarmac was counted. Some of these islands had "complex" vegetation (e.g. trees, bushes, shrubs, ponds and rocks) while others were "simple" (only grass), indicated by the variable `habitatType`,   

Use an appropriate statistical model to explore the relationship between area and species richness. Does this relationship differ depending on habitat complexity?

a) Plot the data to show the relationship between the the number of species and the area of the "island". Colour code the points by whether the habitat type.

b) Fit a suitable statistical model to estimate the statistical relationship between area, habitat type, and species richness. Describe the method and then summarise the results produced by the model as if you were writing a report/thesis.

c) Produce a plot that shows (in addition to the raw data points) the fitted values produced by your model and the uncertainty in those estimates. 

