## Californian bird diversity

1. First import the data. Check that the columns look as they should. (e.g. use `summary` or `str` functions). Tip: use the "Wizard" in RStudio to guide you.

```{r}
birds <- read.csv("CourseData/suburbanBirds.csv")
```


2. What is the mean, minimum, and maximum number of species seen? (there is more than one way to do this)*

```{r}
mean(birds$nSpecies)
min(birds$nSpecies)
max(birds$nSpecies)
range(birds$nSpecies)
summary(birds$nSpecies)
```


3. How old are the youngest and oldest suburbs? (hint: the survey was carried out in 1975, do the math!)

```{r}
1975 - min(birds$Year)
1975 - max(birds$Year)
```

4. Plot the relationship between `Year` and `nSpecies` as a scatterplot using base-R graphics (using the `plot` function).

```{r}
plot(birds$Year, birds$nSpecies)
```

5. The pattern might be easier to see if you could replace `YearBuilt` with suburb age. Create a new vector in your data frame for this variable (e.g. `birds$Age <- 1975 - birds$Year)`). Re-plot your results.

```{r}
birds$Age <- 1975 - birds$Year
plot(birds$Age, birds$nSpecies)
```


6. What do the data show? What might be the mechanisms for the patterns you see? Do they match your expectations?

If you recall that the average species richness pre-development was about 3.5 species, the data show that suburban development is actually good for bird species. This could be surprising, but a possible explanation is that the gardens, parks, trees etc. that come with development represent additional habitats that would not normally be there. Therefore, these areas attract new species.

7. Export your plots and paste them into a Word Document.

You can do this with several methods. My favourite *quick* method is to click the Export button > Copy to Clipboard, resize the plot so it looks nice, then click Copy Plot. Finally, paste into Word with Ctrl (or Cmd) + V.
