## Virtual dice

Let's try the same kind of thing with the roll of (virtual) dice.

Here's how to do one roll of the dice:

```{r, echo=-1, results='asis',cache = TRUE}
set.seed(123)
diceRoll <- 1:6
sample(diceRoll, 1)
```

1) Simulate 10 rolls of the dice, and make a table of results.

```{r, echo=-1,cache = TRUE}
set.seed(123)
result <- sample(diceRoll, 10, replace = TRUE)
table(result)
```

Your table will probably look different to this, because it is a random process. You may notice that some numbers in the table are missing if some numbers were never rolled by our virtual dice. 

2) Now simulate 90 rolls of the dice, and plot the results as a bar plot using `geom_bar` in `ggplot`. Add a horizontal line using `geom_abline` to show the **expected** result based on what you know about probability.


```{r, echo=-1,fig.cap=('Barplot of 90 simulated dice throws'),cache = TRUE}
set.seed(2)
n <- 90
result <- data.frame(result = sample(diceRoll, n, replace = TRUE))

ggplot(result, aes(x = result)) +
  geom_bar() +
  geom_abline(intercept = n / 6, slope = 0)
```

3) Try adjusting the code to simulate dice rolls with small (say, 30) and large (say, 600, or 6000, or 9000) samples. Observe what happens to the proportions, and compare them to the expected value. What does this tell you about the importance of sample size when trying to estimate real phenomena?


You only need to edit the `n <- 90` line in the code above

The main message here is that as sample size increases you are more likely to obtain a good estimate of the true value of the phenomenon.  You may also notice that,  what would be considered a good sample size for the coin flipping (i.e. it recovers the true probability of 0.5 reasonably well) is NOT adequate for getting a good estimate of the probabilities for the dice.

This is because of the different number of possibilities: as the range of possible outcomes increases, the sample size requirements increase. In other words, choosing a good sample size is context-dependent.


```{r, echo= FALSE,fig.cap=('Barplots of 30 and 9000 simulated dice throws'),cache = TRUE}
set.seed(2)
n <- 30
result <- data.frame(result = sample(diceRoll, n, replace = TRUE))

A <- ggplot(result, aes(x = result)) +
  geom_bar() +
  geom_abline(intercept = n / 6, slope = 0) + 
  ggtitle("n = 30")

set.seed(2)
n <- 9000 #Vary this number
result <- data.frame(result = sample(diceRoll, n, replace = TRUE))

B <- ggplot(result, aes(x = result)) +
  geom_bar() +
  geom_abline(intercept = n / 6, slope = 0) + 
  ggtitle("n = 9000")

A + B

```


