## Exercise Solutions: Apple tree crop yield

Import the data (`apples.csv`) and analyse it using the techniques you have learned in the ANOVA lecture, and the previous worksheet, to answer the question "What is the effect of tree spacing on apple yields?"

1. Import and look at the data (e.g. `summary` or `str` or `head`)

```{r}
apples <- read.csv("CourseData/apples.csv")
summary(apples)
```

2. Ensure that the explanatory variable (`spacing`) is defined as a categorical variable (i.e. a "factor", in R-speak). You can use `mutate` and `as.factor` functions for this.

```{r}
apples <- apples %>%
  mutate(spacing = as.factor(spacing))
```



3. Plot the data using `ggplot` (a box plot with (optionally) added jittered points would be good).

```{r, echo = -1,cache=TRUE}
set.seed(42)
ggplot(apples, aes(x = spacing, y = yield)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  ylab("Yield (kg per tree)")
```

4. Fit an ANOVA model using `lm`. 


```{r}
apple_mod <- lm(yield ~ spacing, data = apples)
```

5. Check the model using a diagnostic plot (i.e. using `autoplot` from the `ggfortify` package).

```{r,message = FALSE, warning = FALSE}
library(ggfortify)
autoplot(apple_mod)
```


6. Use `anova` to get the ANOVA summary.

```{r}
anova(apple_mod)
```

7. You should see that there are differences among treatments. But where are those differences? Use `summary` on  your model to find out.

```{r}
summary(apple_mod)
```

8. Use a Tukey test to make all the pairwise comparisons among the treatments.

```{r}
library(agricolae)
HSD.test(apple_mod, "spacing", console = TRUE)
```


9. Write a few sentences that summarise your findings. What biological processes do you think drive the effects that you have detected?

```{r,echo = FALSE}
x <- anova(apple_mod)
```

*There was a significant effect of spacing on apple yields (Figure XX, ANOVA: F = `r formatC(x$"F value"[1], digits = 3, format = "f")`, d.f. = `r x$Df[1]` and `r x$Df[2]`, p = `r formatC(x$"Pr(>F)"[1], digits = 4, format = "f")`). *

Then:  *The pairwise comparisons in the ANOVA model showed that means of the 6 and 10 foot spacing treatment were significantly different (t= 3.244, p = 0.0017), as were those of 6 and 14 (t=4.022, p = 0.0001), but the 10 foot - 14 foot comparison showed no significant difference (t= 0.707, p= 0.4813)^[These values come from the `summary` tables for the ANOVA model, and the releveled ANOVA model].*

Or, more simply: *The 6-10ft and 6-14ft comparisons showed significant differences (Tukey HSD: p<0.05), but the 10-14ft comparison showed no significant difference (Tukey HSD: p>0.05)*

10. Optional. Instead of using a Tukey test, use the alternative "relevel" approach to make the missing comparison.

```{r}
apples2 <- apples %>%
  mutate(spacing = relevel(spacing, ref = "10"))
apple_mod2 <- lm(yield ~ spacing, data = apples2)
summary(apple_mod2)
```


-------
If you get this far, try using the ANOVA approach on one of the previous t-test examples (remember that ANOVA can be used when your single explanatory variable has TWO or more levels). You should notice that the results are the same whether you use the `t.test` function or the ANOVA approach with `lm`.
