## Exercise Solutions: Chirping crickets

1. Import the data

```{r}
chirps <- read.csv("CourseData/chirps.csv")
```

2. Use `mutate` to convert Fahrenheit to Celsius (Google it)

```{r}
chirps <- chirps %>%
  mutate(temperatureC = (temperature - 32) * (5 / 9))
head(chirps)
```

3. Plot the data

```{r}
(A <- ggplot(chirps, aes(x = temperatureC, y = chirps)) +
  geom_point())
```

4. Fit a linear regression model with `lm`

```{r}
chirp_mod <- lm(chirps ~ temperatureC, data = chirps)
```

5. Look at diagnostic plots to evaluate the model

```{r,message = FALSE, warning = FALSE}
library(ggfortify)
autoplot(chirp_mod)
```

6. Use `anova` to figure out if the effect of temperature is statistically significant.

```{r}
anova(chirp_mod)
```

7. Use `summary` to obtain information about the coefficients and $R^2$-value.

```{r}
summary(chirp_mod)
```

8. Summarise the model in words.

*There is a statistically significant association between temperature and chirp frequency (F = `r formatC( anova(chirp_mod)$F[1], digits = 4, format = "f")`, d.f. = `r anova(chirp_mod)$Df[1]`,`r anova(chirp_mod)$Df[2]`, p < 0.001) The equation of the fitted model is: Chirp Freq = `r formatC(summary(chirp_mod)$coef[1,1], digits = 2,format = "f")`($\pm$ `r formatC(summary(chirp_mod)$coef[1,2], digits = 2,format = "f")`) $\times$ Temp + `r formatC(summary(chirp_mod)$coef[2,1], digits = 2,format = "f")`($\pm$ `r formatC(summary(chirp_mod)$coef[2,2], digits = 2,format = "f")`). The model explains `r formatC( summary(chirp_mod)$r.squared*100, digits = 0, format = "f")`% of the variation in chirp frequency ($R^2$ = `r formatC( summary(chirp_mod)$r.squared, digits = 3, format = "f")`).*

9. Add model fit line to the plot.

```{r}
A + geom_smooth(method = "lm")
```

10. Can I use cricket chirp frequency as a kind of thermometer?

Yes, using the equation from the model
