## Exercise Solutions: Fish behaviour

1. Import the data set, `fishPersonality.csv`

```{r}
fishPersonality <- read.csv("CourseData/fishPersonality.csv")
```

2. Plot the data (e.g. as a box plot)

```{r}
ggplot(fishPersonality, aes(x = colouration, y = spawning, fill = personality)) +
  geom_boxplot()
```

3. Fit an ANOVA model using `lm`.

```{r}
mod_A <- lm(spawning ~ personality + colouration + personality:colouration,
  data = fishPersonality
)
```

4. Look at diagnostic plots from this model (`autoplot`)

5. Use `anova` to get an Analysis of Variance summary table, and interpret the results.

```{r}
anova(mod_A)
```


6. Get the coefficient summary (`summary`) and interpret the output.

```{r}
summary(mod_A)
```

7. Do post-hoc Tukey tests (e.g. using `HSD.test` from the `agricolae` package). Interpret the results.

```{r}
library(agricolae)
HSD.test(mod_A, c("personality", "colouration"), console = TRUE)
```


8. Sum up your findings with reference to the initial research questions.


* Homogeneous coloured fish seem to do better than heterogeneous ones overall (from the plot)
* The anova table shows that personality, colouration and the interaction between them are all important variables (p <0.05); Personality seems to be more important than colouration overall (based on p-values)
* The Tukey test table shows that -  (1) Personality is associated with spawning, but only for heterogeneous coloured fish. (2) In the heterogeneous coloured fish the proactive fish spawn significantly more than the reactive ones. (3) In the homogeneous coloured fish there is no significant difference between the personalities.
