## Sexual selection in Hercules beetles

1. What is your null hypothesis?

*Null Hypothesis - There is no difference in the widths of the species.*

2. What is your alternative hypothesis?

*Alternative Hypothesis - Males have larger widths than females.*

3. Import the data.
```{r}
hercules <- read.csv("CourseData/herculesBeetle.csv")
```

4. Calculate the mean for each sex (either using `tapply` or using `dplyr` tools)

```{r}
# With dplyr
hercules %>%
  group_by(sex) %>%
  summarise(mean = mean(width))

# with tapply
tapply(hercules$width, hercules$sex, mean)
```

5. Plot the data as a histogram.

```{r,message=FALSE}
# Let's look at the male and female data
(plot1 <- ggplot(hercules, aes(x = width, fill = sex)) +
  geom_histogram(position = "identity", alpha = .7))
```

6. Add vertical lines to the plot to indicate the mean values.

```{r,message=FALSE}
hercules %>%
  group_by(sex) %>%
  summarise(mean = mean(width)) %>%
  pull(mean) -> meanVals

plot1 + geom_vline(xintercept = meanVals, colour = c("red", "blue"))
```

7. Now calculate the difference between the mean values using `dplyr` tools, or `tapply`.

```{r}

# with dplyr
hercules %>%
  group_by(sex) %>%
  summarise(mean = mean(width)) %>%
  pull(mean) %>%
  diff() -> observedDiff
observedDiff

# with tapply
diff(as.vector(tapply(hercules$width, hercules$sex, mean)))
```

8. Use `sample` to randomise the sex column of the data, and recalculate the difference between the mean.

```{r}
# with dplyr
hercules %>%
  mutate(sex = sample(sex)) %>%
  group_by(sex) %>%
  summarise(mean = mean(width)) %>%
  pull(mean) %>%
  diff()

# with tapply
diff(as.vector(tapply(hercules$width, sample(hercules$sex), mean)))
```

9. Use `replicate` to repeat this 10 times (to ensure that you code works).

```{r}
# with dplyr
replicate(
  10,
  hercules %>%
    mutate(sex = sample(sex)) %>%
    group_by(sex) %>%
    summarise(mean = mean(width)) %>%
    pull(mean) %>%
    diff()
)

# with tapply
replicate(
  10,
  diff(as.vector(tapply(
    hercules$width,
    sample(hercules$sex), mean
  )))
)
```

10. When your code is working, use `replicate` again, but this time with 1000 replicates and pass the results into a data frame.

```{r}
# with dplyr
diffs <- data.frame(
  diffs =
    replicate(
      1000,
      hercules %>%
        mutate(sex = sample(sex)) %>%
        group_by(sex) %>%
        summarise(mean = mean(width)) %>%
        pull(mean) %>%
        diff()
    )
)

# with tapply
diffs <- data.frame(
  diffs =
    replicate(
      1000,
      diff(as.vector(
        tapply(hercules$width, sample(hercules$sex), mean)
      ))
    )
)
```

11. Use `ggplot` to plot the null distribution you have just created, and add the observed difference.

```{r,message=FALSE}
ggplot(diffs, aes(x = diffs)) +
  geom_histogram() +
  geom_vline(xintercept = observedDiff)
```

12. Obtain the p-value for the hypothesis test described above. (1) how many of the observed differences are greater than or equal to the shuffled differences in the null distribution. (2) what is this expressed as a proportion of the number of replicates.

```{r}
sum(observedDiff >= diffs$diffs)
sum(observedDiff >= diffs$diffs) / 1000
```

13. Summarise your result as in a report. Describe the method, followed by the result and conclusion.

*"I used a randomisation test to estimate the significance of the observed difference of `r format(observedDiff,digits= 3,nsmall=3)` (mean values: female=`r format(meanVals[2],digits= 3,nsmall=3)`;  male = `r format(meanVals[1],digits= 3,nsmall=3)`) in mean widths of the sexes. To do this I generated a null distribution of differences between sexes using 1000 replicates. I found that only `r sum(observedDiff>=diffs$diffs)` of the differences in the null distribution were as extreme as the observed difference. Thus the p-value is `r sum(observedDiff>=diffs$diffs)/1000`: I therefore reject the null hypothesis that there is no difference between the sexes and accept the alternative hypothesis that males are significantly larger than females."*
