## Sex differences in fine motor skills

Some people have suggested that there might be sex differences in fine motor skills in humans. Use the data collected on the class to address this topic using t-tests. The relevant data set is called `classData.csv`, and the columns of interest are `Gender` and `Precision`.

Carry out a two-sample t-test. 

1) Plot the data (e.g. with a box plot, or histogram)

```{r,echo = -1,cache=TRUE}
set.seed(12)
classData <- read.csv("CourseData/classData.csv") %>% 
  filter(Gender %in% c("Male", "Female"))

ggplot(classData, aes(x = Gender, y = Precision)) +
  geom_boxplot() +
  geom_jitter(width = 0.2)
```


2) Formulate null and alternative hypotheses. 

Null hypotheses - the differences in precision between male and female are due to chance alone.
Alternative hypothesis - there is a significant difference between male and female precision scores.

3) Use the `t.test` function to do the test.


```{r}
t.test(Precision ~ Gender, data = classData)
```

4) Write a sentence or two describing the results.
```{r, echo = FALSE}
x <- t.test(Precision ~ Gender, data = classData)
```
There was no significant difference in mean precision between the two genders (t-test: t = `r format(x$statistic,nsmall = 2,digits = 3)`, df = `r format(x$parameter,nsmall = 2,digits = 3)`, p = `r format(x$p.value,nsmall = 2,digits = 3)`). The 95% confidence interval for the difference between genders overlapped 0 (95% CI = `r sprintf(x$conf.int[1], fmt = '%#.3f')`-`r sprintf(x$conf.int[2], fmt = '%#.3f')`). I therefore fail to reject the null hypothesis that the observed differences are due to chance alone.

## Therapy for anorexia

A study was carried out looking at the effect of cognitive behavioural therapy on weight of people with anorexia. Weight was measured in week 1 and again in week 8. Use a paired t-test to assess whether the treatment is effective.

The data is called `anorexiaCBT.csv`

The data are in "wide format". You may wish to convert it to "long format" depending on how you use the data. You can do that with the `pivot_longer` function, which rearranges the data:


1) Plot the data (e.g. with an interaction plot like Figure \@ref(fig:toolData))

```{r,echo=TRUE,message=FALSE}
anorexiaCBT <- read.csv("CourseData/anorexiaCBT.csv",
  header = TRUE
)

anorexiaCBT_long <- anorexiaCBT %>%
  pivot_longer(
    cols = starts_with("Week"), names_to = "time",
    values_to = "weight"
  )
```

```{r}
ggplot(anorexiaCBT_long, aes(
  x = time, y = weight,
  group = Subject
)) +
  geom_line() +
  geom_point() +
  xlab("Time") +
  ylab("Weight (kg)")
```



2) Formulate a null and alternative hypothesis.

Null = The difference in weight between the two times is no different than random chance.
Alternative Hypothesis = There is a significant change in weight between the two time points.

3) Use `t.test` to conduct a *paired* t-test.

The method here depends on whether you use the "long" data or not:

```{r}
t.test(anorexiaCBT$Week01, anorexiaCBT$Week08, paired = TRUE)
```

or

```{r}
anorexiaCBT_long <- anorexiaCBT_long %>%
  arrange(Subject, time)

t.test(weight ~ time, data = anorexiaCBT_long, paired = TRUE)
```


4) Write a couple of sentences to report your result.

```{r}
x <- t.test(anorexiaCBT$Week01, anorexiaCBT$Week08, paired = TRUE)
```

There was a significant difference in weight of `r sprintf(x$estimate, fmt= '%#.3f')`kg between week 1 and week 8 (t.test: t = `r format(x$statistic,nsmall = 2,digits = 3)`, df = `r format(x$parameter,nsmall = 2,digits = 3)`, p = `r format(x$p.value,nsmall = 2,digits = 3)`). The 95% confidence interval for the difference between weeks was between `r sprintf(x$conf.int[1], fmt = '%#.3f')` and `r sprintf(x$conf.int[2], fmt = '%#.3f')`. Therefore, I reject the null hypotheses, that the difference is due to chance alone, and accept the alternative hypothesis.

## Compare t-tests with randomisation tests

```{block, type="do-something"}
Try re-fitting some of these tests as randomisation tests (or analyse the randomisation test data using `t.test`). Do they give approximately the same results? 

Then try answering the question - "*are people who prefer dogs taller than those who prefer cats?* " using the `classData.csv`. Can you think of any problems with this analysis?
```

The problem with the analysis is that it is "confounded". That is to say, gender is correlated with height, so you would not be sure whether any difference you found would be due to height, or gender.
